---
title: "Analysis of the Complete Block Design"
author: Prof Randi Garcia, SDS 290
date: October 9, 2024
format: 
  revealjs:
    scrollable: true
    theme: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)
```

## Announcements

-   Grades posted for HW4
-   HW5 posted on Friday
-   MP1 pre-approval due Friday 11:55p
-   Office hours (Bass 412)
    -   Tuesday 10:30a - 11:30a
    -   Friday office hours cancelled (again)
    -   Thursday 4p-5p or by appointment

## MP1 Pre-Approval

Your completed Qualtrics experiments are due for Pre-Approval by **Fri Oct 11 11:59p**.

1.  Export your survey as a Word doc. Include the survey flow and any skip logic you used.
2.  Upload the Word doc to Moodle.

I should be able to see any images you used or links to media. I'll be paying particular attention to your consent form. It's not unusual for me to ask for revisions before you launch your study.

## Next steps:

I will give you approval or revisions via Slack direct messages. Watch out for those!

Once you have completed any revisions and receive approval from me you can complete the following **checklist** and then share your anonymous survey link.

1.  Proof read for errors.
2.  Test (preview) your survey multiple times, creating fake data.
3.  [Download the data](https://www.qualtrics.com/support/survey-platform/data-and-analysis-module/data/download-data/export-data-overview/) and make sure it makes sense.
4.  Zero out the [randomization counts](https://www.qualtrics.com/support/survey-platform/survey-module/survey-flow/standard-elements/randomizer/#EditingCounts) in the survey flow.
5.  [Delete](https://www.qualtrics.com/support/survey-platform/data-and-analysis-module/data/recorded-responses/#DeletingResponses) all of the fake responses.
6.  [Upgrade](https://www.qualtrics.com/support/survey-platform/sp-administration/coupon-codes/#UpgradingasanExistingUser) your Qualtrics account with the code **SDS290Fall2024**.
7.  Activate your survey.
8.  Get the [anonymous survey link](https://www.qualtrics.com/support/survey-platform/distributions-module/distributions-overview/)!

## Agenda

1.  Complete Block Design

## Inappropriate Insects

Modern zoos try to reproduce natural habitats in their exhibits as much as possible. They try to use appropriate plants, but these plants can be infested with inappropriate insects. Cycads (plants that look vaguely like palms) can be infected with mealybugs, and the zoo wishes to test three treatments: 1) water, 2) horticultural oil, and 3) fungal spores in water. Nine infested cycads are taken to the testing area. Three branches are randomly selected from each tree, and 3 cm by 3 cm patches are marked on each branch. The number of mealybugs on the patch is counted. The three treatments then get randomly assigned to the three branches for each tree. After three days the mealybugs are counted again. The change in number of mealybugs is computed ($before-after$).

## Inappropriate Insects

```{r, echo=FALSE}
library(tidyr)
library(dplyr)
library(knitr)
library(tidyverse)

tree <- c(rep("tree1",3), rep("tree2",3), rep("tree3", 3),rep("tree4", 3), rep("tree5", 3))
treatment <- rep(c("water", "spores", "oil"), 5)
bugs_change <- c(-9, -4, 4, 18, 29, 29, 10, 4, 14, 9, -2, 14, -6, 11, 7)

mealybugs <- data.frame(tree, treatment, bugs_change)

mealybugs %>%
  spread(tree, bugs_change) %>%
  kable()
```

Draw the factor diagram.

## MORE Design Terminology

-   The two factors *Trees* and *Treatment* are **crossed**. This means that we collect data on all combinations of *Trees* and *Treatment*. A **cell** is a particular combination.
-   **Main effect** refers to the effect for a single factor. For a block design there are two sets of main effects, for blocks and for treatments (e.g., *Trees* and *Treatment*).
-   The two sets of main effects are called **additive** if each treatment adds or subtracts the same amount to the response variable for all blocks.
-   A **factor of interest** is *Treatment*, which is directly related to the goal of the study. A **nuisance factor** is *Trees* which is included as a factor only because it is a source of variation and its inclusion will reduce the sum of squares for error (residuals).
<!-- -   Typically, factors of interest are **fixed**, which means their effects are an unknown constant. Nuisance factors are **random**, which means their effects behave according to chance (the levels themselves are samples from a population). -->

## Side-by-side dotplot

We add our blocking factor as `color` and also as `group`.

```{r}
ggplot(mealybugs, aes(x = treatment, y = bugs_change, color = tree, group = tree)) + 
  geom_line()
```

We can see that Tree 2's infestation was very responsive to the treatments whereas Tree 1's was not.

<!-- ## Side-by-side dotplot (adjusted for effect of blocks) -->

<!-- ```{r, include=FALSE} -->

<!-- mealybugs <- mealybugs %>% -->

<!--   ungroup() %>% -->

<!--   mutate(grand = mean(bugs_change)) %>% -->

<!--   group_by(tree) %>% -->

<!--   mutate(tree_mean = mean(bugs_change), -->

<!--          tree_effect = tree_mean - grand, -->

<!--          bugs_change_adj = bugs_change - tree_effect) %>% -->

<!--   ungroup() %>% -->

<!--   group_by(treatment) %>% -->

<!--   mutate(treatment_effect = mean(bugs_change)) -->

<!-- ``` -->

<!-- ```{r} -->

<!-- qplot(x = treatment, bugs_change_adj, data = mealybugs) -->

<!-- ``` -->

## Formal ANOVA for Complete Block Design

$${y}_{ij}={\mu}+{\tau}_{i}+{\beta}_{j}+{e}_{ij}$$

|    Source |                                         SS                                          |      df      |             MS              |              F              |
|--------------:|:-------------:|:-------------:|:-------------:|:-------------:|
| Treatment |                  $\sum_{i=1}^{a}b(\bar{y}_{i.}-\bar{y}_{..})^{2}$                   |    $a-1$     | $\frac{{SS}_{T}}{{df}_{T}}$ | $\frac{{MS}_{T}}{{MS}_{E}}$ |
|    Blocks |                  $\sum_{j=1}^{b}a(\bar{y}_{.j}-\bar{y}_{..})^{2}$                   |    $b-1$     | $\frac{{SS}_{B}}{{df}_{B}}$ | $\frac{{MS}_{B}}{{MS}_{E}}$ |
|     Error | $\sum_{i=1}^{a}\sum_{j=1}^{b}({y}_{ij}-\bar{y}_{i.}-\bar{y}_{.j}+\bar{y}_{..})^{2}$ | $(a-1)(b-1)$ | $\frac{{SS}_{E}}{{df}_{E}}$ |                             |

## Data Analysis Structure

```{r, include=FALSE}
rm(mealybugs)
tree <- c(rep("tree1",3), rep("tree2",3), rep("tree3", 3),rep("tree4", 3), rep("tree5", 3))
treatment <- rep(c("water", "spores", "oil"), 5)
bugs_change <- c(-9, -4, 4, 18, 29, 29, 10, 4, 14, 9, -2, 14, -6, 11, 7)
mealybugs <- data.frame(tree, treatment, bugs_change)
```

```{r}
mealybugs
```

## Formal ANOVA

```{r}
mod <- lm(bugs_change ~ treatment + tree, data = mealybugs)

anova(mod)
```

Write a sentence interpreting the results of the F-test for treatment and a sentence interpreting the F-test for tree in the context of the problem.

## Formal ANOVA

```{r}
mod <- lm(bugs_change ~ treatment + tree, data = mealybugs)

anova(mod)
```

There are no statistically significant differences in the reduction in mealy bugs between the three treatment conditions, $F(2, 8) = 3.00$, $p = .107$. There are significant differences in the reduction in mealy bugs across trees, however, $F(4, 8) = 9.04$, $p = .005$. That is, some trees improved more than other trees.

## Informal Analysis Structure

```{r, }
mealybugs_wide <- mealybugs %>%
  spread(treatment, bugs_change)

mealybugs_wide
```

## How to check assumptions

-   C. **Constant effects** -- *think* about whether it is reasonable.

-   A. **Additive effects** -- check Anscombe block plots.

-   S. **Same standard deviations** -- is the biggest SD less than two times as large as the smallest? check residual versus fitted plot: *does the plot thicken*?

-   I. **Independent residuals** -- *think* about whether it is reasonable.

-   N. **Normally distributed residuals** -- construct a histogram or normal probability plot of residuals.

-   Z. **Zero mean residuals** -- construct a histogram or normal probability plot of residuals.

## Anscombe Block Plots

-   Scatterplots of two levels (e.g., level 1 vs. level 2) of the factor of interest.
-   Used for
    -   exploring the data, and
    -   assessing the **additivity (A)** condition.
-   If we compare a spore observation to an oil observation from the same tree, the only difference should be the difference in treatment effects.
-   Thus, the line-of-best fit should have a slope of approximately 1. 

## Informal Analysis Structure

```{r}
mealybugs_wide <- mealybugs %>%
  pivot_wider(names_from = treatment, values_from = bugs_change)

mealybugs_wide
```

## Anscombe Block Plots

```{r}
qplot(x = spores, y = oil, data = mealybugs_wide) +
  geom_abline(intercept = 13.6-7.6, slope = 1, color = "blue", linetype = 2) + 
  geom_smooth(method = "lm", se = 0, color = "orange")
```

## Anscombe Block Plots

```{r}
qplot(x = spores, y = water, data = mealybugs_wide) +
  geom_abline(intercept = 4.4 - 13.6, slope = 1, color = "blue", linetype = 2) + 
  geom_smooth(method = "lm", se = 0, color = "orange")
```

## Anscombe Block Plots

```{r}
qplot(x = oil, y = water, data = mealybugs_wide) +
  geom_abline(intercept = 4.4 - 13.6, slope = 1, color = "blue", linetype = 2) + 
  geom_smooth(method = "lm", se = 0, color = "orange")
```

## Assessing S Condition

```{r}
mealybugs %>%
  group_by(treatment) %>%
  summarize(m = mean(bugs_change),
            sd = sd(bugs_change))

mealybugs %>%
  group_by(treatment) %>%
  summarize(m = mean(bugs_change),
            sd = sd(bugs_change)) %>%
  summarize(max(sd)/min(sd)) #calculating using min and max function
```

## Assessing S Condition

$$\hat{{y}}_{ij}={\mu}+{\tau}_{i}+{\beta}_{j}$$

-   Where $\hat{{y}}_{ij}$ are the **fitted values**, that is, everything but the ticket at the end of the assembly line.

```{r}
plot(mod, which = 1)
```

If the **plot thickens**, that is, has a patterning that looks like a funnel, then the S condition is not satisfied.

## Assessing N Condition

```{r}
plot(mod, which = 2)
```

We're looking for residuals to be on the line. If so, then we can say they are normally distributed.

## Assessing the Z Condition

```{r}
qplot(mod$residuals, bins = 6)
```

If the histogram centered at zero? Then the Z condition is satisfied.

## Sleeping Shrews

Assess the CA-SINZ conditions for the `SleepingShrews` dataset from example 6.7b in your textbook.

```{r}
library(Stat2Data)
data("SleepingShrews")

#You'll need the WIDE dataset for the Anscombe block plots
SleepingShrews_wide <- SleepingShrews %>%
  select(-ID) %>%
  pivot_wider(names_from = Phase, values_from = Rate)
```

## Testing Condition for Sleeping Shrews Data

```{r, eval = FALSE}
qplot(x = DSW, y = LSW, data = SleepingShrews_wide) +
  geom_abline(intercept = 2, slope = 1, color = "blue", linetype = 2) +   
  geom_smooth(method = "lm", se = 0, color = "orange") #A condition

qplot(x = DSW, y = REM, data = SleepingShrews_wide) +
  geom_abline(intercept = 2, slope = 1, color = "blue", linetype = 2) +   
  geom_smooth(method = "lm", se = 0, color = "orange") #A condition

qplot(x = LSW, y = REM, data = SleepingShrews_wide) +
  geom_abline(intercept = -2, slope = 1, color = "blue", linetype = 2) +   
  geom_smooth(method = "lm", se = 0, color = "orange") #A condition

SleepingShrews %>%
  group_by(Phase) %>%
  summarize(m = mean(Rate),
            sd = sd(Rate)) %>%
  summarize(max(sd)/min(sd)) #S condition

mod <- lm(Rate ~ Phase + Shrew, data = SleepingShrews)
plot(mod, which = 1) #S condition
plot(mod, which = 2) #N condition
qplot(mod$residuals, bins = 6) #N and Z conditions
```
